Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples utilisables par Isabelle et Coq'' (CUIC). }

### 2015-06-04 tex/body.tex
fixed typo
more

\newcommand{\sfuns}[1]{#1^\mathrm{f}}
\newcommand{\sfundefs}[1]{#1^\mathrm{dfn}}
%\newcommand{\sfunndefs}[1]{#1^\mathrm{nf}}
\newcommand{\typeint}{\ty{Int}}

@@

Our setting is a monomorphic (or many-sorted) first-order logic
like the one defined by SMT-LIB \cite{smtlib25}.
A \emph{signature} $\Sigma$ consists of
a set $\stypes{\Sigma}$ of %ct type symbols
first-order types
%
%\ct{What the rationale for using the letter $y$ in $\stypes{\Sigma}$?}
%\jb{y as in t\textbf{y}pe. changed to `ty'}
(or sorts) and a set $\sfuns{\Sigma}$ of function symbols over these types.
We assume that signatures always contain a Boolean type $\Bool$ and constants
$\ltrue, \lfalse : \Bool$ for truth and falsity,
an infix equality predicate ${\teq} : \tau \times \tau \to \Bool$
for each $\tau \in \stypes{\Sigma}$,
standard Boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and an if--then--else function symbol
$\lite : \Bool \times \tau \times \tau \rightarrow \tau$
for each $\tau \in \stypes{\Sigma}$.
Formulas are terms of type $\Bool$.
We write $t\typ{\tau}$ to denote a metavariable ranging over terms of
type~$\tau$ and $\terms( t )$ to denote the set of subterms in $t$.
Given a term $u$, we write $u[\vec t/\vec x]$ to denote the result of replacing
all occurrences of $\vec x$ with $\vec t$ in $u$.
\begin{longv}When applied to terms, the symbol $=$ denotes syntactic equality.\end{longv}

@@

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each type $\tau \in \stypes{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in~$\I$,
and each function symbol $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in
$\sfuns{\Sigma}$
to a total function $\con{f}^\I : \tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where
$\Sigma$ is a signature and $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$.
A $\Sigma$-formula $\varphi$ is
\emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$\begin{longv};
otherwise, it is \emph{$T$-unsatisfiable}\end{longv}.
A formula $\varphi$ \emph{$T$-entails} $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Two formulas $\varphi$ and $\psi$ are \emph{$T$-equivalent} 
if each $T$-entails the other.
Given a signature $\Sigma$,
the \emph{theory of equality with uninterpreted functions} \euf
consists of the set of all $\Sigma$-interpretations.
Informally, we refer to the type and function symbols in this theory as \emph{uninterpreted}.
%\ct{This is both ambiguous and insufficient.
%Since this is SMT we should fix some background theory $T$,
%then extend its signature with additional symbols and call \emph{those} uninterpreted.
%}

@@

\newcommand\closurefmla{\psi \models_T \bigwedge\nolimits_{\,i=1}^{n} \{ \psi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \psi ) \}}

@@

The Leon system~\cite{blanc2013overview} implements a procedure that can
produce both proofs and counter\-examples for properties of terminating functions
written in a subset of Scala. Leon is based on an SMT solver. It avoids
quantifiers altogether by unfolding recursive definitions up to a certain
depth, which is increased on a per-need basis.
Our translation~$\conv$ works in an analogous manner, 
where instead the SMT solver is invoked only once 
and where quantifier instantiation amounts to function unfolding.
\begin{longv}It would be
worth investigating how existing approaches for function
unfolding can inform approaches for dedicated quantifier instantiation
techniques for function definitions, and vice versa.\end{longv}

@@

%\begin{nottinyv}
{%\footnotesize
\def\ackname{Acknowledgment}
\paragraph{%\footnotesize
\ackname.}
Viktor Kuncak and Stephan Merz have made this work possible. We would also like
to thank Damien Busato-Gaston and Emmanouil Koukoutos for providing the initial
set of benchmarks used in the evaluation.
}
%\end{nottinyv}

>>>>

\newcommand{\svars}[1]{#1^\mathrm{v}}
\newcommand{\sfuns}[1]{#1^\mathrm{f}}
\newcommand{\sfundefs}[1]{#1^\mathrm{dfn}}
%\newcommand{\sfunndefs}[1]{#1^\mathrm{nf}}
\newcommand{\typeint}{\ty{Int}}

@@

Our setting is a monomorphic (or many-sorted) first-order logic
like the one defined by SMT-LIB \cite{smtlib25}.
A \emph{signature} $\Sigma$ consists of
a set $\stypes{\Sigma}$ of %ct type symbols
first-order types
%
%\ct{What the rationale for using the letter $y$ in $\stypes{\Sigma}$?}
%\jb{y as in t\textbf{y}pe. changed to `ty'}
(or sorts) and a set $\sfuns{\Sigma}$ of function symbols over these types.
We assume that signatures always contain a Boolean type $\Bool$ and constants
$\ltrue, \lfalse : \Bool$ for truth and falsity,
an infix equality predicate ${\teq} : \tau \times \tau \to \Bool$
for each $\tau \in \stypes{\Sigma}$,
standard Boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and an if--then--else function symbol
$\lite : \Bool \times \tau \times \tau \rightarrow \tau$
for each $\tau \in \stypes{\Sigma}$.
%
For each $\tau \in \stypes{\Sigma}$,
we fix an infinite set $\svars{\Sigma}_\tau$ of \emph{variables of type $\tau$} and
define $\svars{\Sigma}$ as $\bigcup_{\tau \in \stypes{\Sigma}} \svars{\Sigma}_\tau$.
$\Sigma$-terms are built as usual over functions symbols in $\Sigma$ and variables in $\svars{\Sigma}$.
%
Formulas are terms of type $\Bool$.
We write $t\typ{\tau}$ to denote a metavariable ranging over terms of
type~$\tau$ and $\terms( t )$ to denote the set of subterms in $t$.
Given a term $u$, we write $u[\vec t/\vec x]$ to denote the result of replacing
all occurrences of $\vec x$ with $\vec t$ in $u$.
\begin{longv}When applied to terms, the symbol $=$ denotes syntactic equality.\end{longv}

@@

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each type $\tau \in \stypes{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in~$\I$,
each function symbol $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in $\sfuns{\Sigma}$
to a total function $\con{f}^\I : \tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$,
and each variable $x:\tau$ of $\svars{\Sigma}$ to an element of $\tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where
$\Sigma$ is a signature and $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$.
A $\Sigma$-formula $\varphi$ is
\emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$\begin{longv};
otherwise, it is \emph{$T$-unsatisfiable}\end{longv}.
A formula $\varphi$ \emph{$T$-entails} $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Two formulas $\varphi$ and $\psi$ are \emph{$T$-equivalent} 
if each $T$-entails the other.
Given a signature $\Sigma$,
the \emph{theory of equality with uninterpreted functions} \euf
consists of the set of all $\Sigma$-interpretations.
Informally, we refer to the type and function symbols in this theory as \emph{uninterpreted}.
%\ct{This is both ambiguous and insufficient.
%Since this is SMT we should fix some background theory $T$,
%then extend its signature with additional symbols and call \emph{those} uninterpreted.
%}

@@

\newcommand\closurefmla{\psi \models_T \bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \psi ) \}}

@@

The Leon system~\cite{blanc2013overview} implements a procedure that can
produce both proofs and counter\-examples for properties of terminating functions
written in a subset of Scala. Leon is based on an SMT solver. It avoids
quantifiers altogether by unfolding recursive definitions up to a certain
depth\begin{longv}, which is increased on a per-need basis\end{longv}.
Our translation~$\conv$ works in an analogous manner, 
where instead the SMT solver is invoked only once 
and where quantifier instantiation amounts to function unfolding.
\begin{longv}It would be
worth investigating how existing approaches for function
unfolding can inform approaches for dedicated quantifier instantiation
techniques for function definitions, and vice versa.\end{longv}

@@

%\begin{nottinyv}
{%\footnotesize
\def\ackname{Acknowledgment}
\paragraph{%\footnotesize
\ackname.}
Viktor Kuncak and Stephan Merz have made this work possible. We would also like
to thank Damien Busato-Gaston and Emmanouil Koukoutos for providing the initial
set of benchmarks used in the evaluation, and the anonymous reviewers for their
suggestions and comments.
}
%\end{nottinyv}

### 2015-06-04 tex/camera_ready/paper.tex
more

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each type $\tau \in \stypes{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in~$\I$,
 each function symbol $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in $\sfuns{\Sigma}$
to a total function $\con{f}^\I : \tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$,
and each variable $x:\tau$ of $\svars{\Sigma}$ as an element of $\tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where
$\Sigma$ is a signature and $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$, closed under variable reassignment
(i.e., for every $I \in \Mo$, every $\Sigma$-interpretation that differs
from $I$ only on the variables of $\svars \Sigma$ is also in $\Mo$).
A $\Sigma$-formula $\varphi$ is \emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$\begin{longv};
otherwise, it is \emph{$T$-unsatisfiable}\end{longv}.
A formula $\varphi$ \emph{$T$-entails} $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Two formulas $\varphi$ and $\psi$ are \emph{$T$-equivalent} 
if each $T$-entails the other.
%ct changed the following because it is not adequate for our purposes
%
%Given a signature $\Sigma$,
%the \emph{theory of equality with uninterpreted functions} \euf
%consists of the set of all $\Sigma$-interpretations.
%Informally, we refer to the type and function symbols in this theory as \emph{uninterpreted}.
%%\ct{This is both ambiguous and insufficient.
%%Since this is SMT we should fix some background theory $T$,
%%then extend its signature with additional symbols and call \emph{those} uninterpreted.
%%}
If $T_1 = (\Sigma_1, \Mo_1)$ is a theory and $\Sigma_2$ is a signature 
with $\sfuns{\Sigma_1} \cap \sfuns{\Sigma_2} = \emptyset$,
the \emph{extension of $T_1$ to $\Sigma_2$} is the theory $T = (\Sigma, \Mo)$ where 
$\sfuns{\Sigma} = \sfuns{\Sigma_1} \cup \sfuns{\Sigma_2}$,
$\stypes{\Sigma} = \stypes{\Sigma_1} \cup \stypes{\Sigma_2}$,
and $\Mo$ is the set of all $\Sigma$-interpretations $\I$
whose $\Sigma_1$-reduct is a model of $T_1$.
We refer to the symbols of $\Sigma_2$ that are not in $\Sigma_1$ as \emph{uninterpreted}.
For the rest of the paper, \relax{we fix a theory $T = (\Sigma, \Mo)$ 
with uninterpreted symbols} constructed as above.

>>>>

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each type $\tau \in \stypes{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in~$\I$,
 each function symbol $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in $\sfuns{\Sigma}$
to a total function $\con{f}^\I : \tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$,
and each variable $x:\tau$ of $\svars{\Sigma}$ to an element of $\tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where
$\Sigma$ is a signature and $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$, closed under variable reassignment
(i.e., for every $I \in \Mo$, every $\Sigma$-interpretation that differs
from $I$ only on the variables of $\svars \Sigma$ is also in $\Mo$).
A $\Sigma$-formula $\varphi$ is \emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$\begin{longv};
otherwise, it is \emph{$T$-unsatisfiable}\end{longv}.
A formula $\varphi$ \emph{$T$-entails} $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Two formulas $\varphi$ and $\psi$ are \emph{$T$-equivalent} 
if each $T$-entails the other.
%ct changed the following because it is not adequate for our purposes
%
%Given a signature $\Sigma$,
%the \emph{theory of equality with uninterpreted functions} \euf
%consists of the set of all $\Sigma$-interpretations.
%Informally, we refer to the type and function symbols in this theory as \emph{uninterpreted}.
%%\ct{This is both ambiguous and insufficient.
%%Since this is SMT we should fix some background theory $T$,
%%then extend its signature with additional symbols and call \emph{those} uninterpreted.
%%}
If $T_1 = (\Sigma_1, \Mo_1)$ is a theory and $\Sigma_2$ is a signature 
with $\sfuns{\Sigma_1} \cap \sfuns{\Sigma_2} = \emptyset$,
the \emph{extension of $T_1$ to $\Sigma_2$} is the theory $T = (\Sigma, \Mo)$ where 
$\sfuns{\Sigma} = \sfuns{\Sigma_1} \cup \sfuns{\Sigma_2}$,
$\stypes{\Sigma} = \stypes{\Sigma_1} \cup \stypes{\Sigma_2}$,
and $\Mo$ is the set of all $\Sigma$-interpretations $\I$
whose $\Sigma_1$-reduct is a model of $T_1$.
We refer to the symbols of $\Sigma_2$ that are not in $\Sigma_1$ as \emph{uninterpreted}.
For the rest of the paper, \relax{we fix a theory $T = (\Sigma, \Mo)$ 
with uninterpreted symbols} constructed as above.
